---
title: "RNA_seq_DEA"
author: "Maryam"
date: "2025-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::#"DESeq2",force = TRUE)
BiocManager::install("edgeR",force = TRUE)

```

```{r}
library(edgeR)
library(DESeq2)
```
now we are reading the data file 
```{r}
Coinfection.targets<-read.delim("./data/fileDesc.txt")
rownames(Coinfection.targets)<-c("Sl1","Sl2","Sl3","Ctr1","Ctr2","Ctr3")
```


```{r}
dim(Coinfection.orig)
```
```{r}
Coinfection.orig <- readDGE(Coinfection.targets, header=F)
```


```{r}
head(Coinfection.orig)
```
```{r}
Coinfection.rawCount <- Coinfection.orig$count
dim(Coinfection.rawCount)
```
```{r}
```


```{r}
sampletype <- factor(c(rep("Sl",3), rep("Ctr", 3)))
```

#Build meta data frame
```{r}
meta <- data.frame(sampletype, row.names = colnames(Coinfection.orig$count))
```

#check the column name of counts dataframe
```{r}
colnames(Coinfection.orig$count)
```
#check the rowname of meta dataframe
```{r}
rownames(meta)
```
#Check that sample names match in both files
```{r}
all(colnames(Coinfection.orig$count) %in% rownames(meta))
```
#-----------------------------------------
# Create DESeqDataset object
#----------------------------------------
#load the package DESeq2 Create DESEq2 dataset dds
```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(Coinfection.orig, colData = meta, design = ~ sampletype)
```
```{r}

```

```{r}
head(counts(dds))
```
#Generate size factors
```{r}
dds <- estimateSizeFactors(dds)
```

```{r}
sizeFactors(dds)
```
#Generate the normalized counts

```{r}
normalized_counts <- counts(dds, normalized=TRUE)
```


```{r}
write.csv(normalized_counts, file="./results/coinfection_normalized_counts_DESeq2.csv")
```


```{r}
rld <- rlog(dds, blind=TRUE)
```

```{r}
plotPCA(rld, intgroup="sampletype")
```
```{r}
pdf("./results/PlotPCA_dds.pdf")
plotPCA(rld, intgroup="sampletype")
```

```{r}
pdf("./results/PlotPCA_dds.pdf")
plotPCA(rld, intgroup="sampletype")
dev.off()
```
#Hierarchical Clustering Heatmap

```{r}
rld_mat <- assay(rld)
```

```{r}
rld_cor <- cor(rld_mat) 
```

```{r}
head(rld_cor)
```
```{r}
head(meta)
```



```{r}
install.packages("pheatmap")
```


```{r}
library(pheatmap)
pheatmap(rld_cor, annotation = meta)
```


```{r}
heat.colors <- RColorBrewer::brewer.pal(6, "Blues")
pheatmap(rld_cor, annotation = meta, color = heat.colors, border_color=NA, fontsize = 10, 
        fontsize_row = 10, height=20)
```


```{r}
pdf("./results/PlotHeatmap_dds.pdf")
heat.colors <- RColorBrewer::brewer.pal(6, "Blues")
pheatmap(rld_cor, annotation = meta, color = heat.colors, border_color=NA, fontsize = 10, 
        fontsize_row = 10, height=20)
```

#Differential expression analysis (DEA) using EdgeR

```{r}
library(edgeR)
options(digits=3)
```

```{r}
```


```{r}
infection.targets
```

```{r}

benefit.targets<-read.delim("./data/fileDesc.txt")
rownames(beneficial.inoculum)<-c("Sl","Sl","Sl","Ctr1","Ctr2","Ctr3")

```

```{r}
getwd()
```

```{r}
library(edgeR)
options(digits=3)
```

```{r}
benefit.targets<-read.delim("./data/fileDesc.txt")
```

```{r}
benefit.targets
```

```{r}
rownames(benefit.targets)<-c("Sl1","Sl2","Sl3","Ctr1","Ctr2","Ctr3")
```

```{r}
benefit.targets
```

```{r}
benefitialinnoculum<- readDGE(benefit.targets, header=F)
```

```{r}
dim(benefitialinnoculum)
```

```{r}
head(benefitialinnoculum)
```

```{r}
benefitialinnoculum.rawCount <- benefitialinnoculum$count
head(benefitialinnoculum.rawCount)
```
```{r}
```


```{r}
install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

```{r}
ggplot(benefitialinnoculum.rawCount) +
  geom_histogram(aes(x = Sl1), stat = "bin", bins = 200) +
  xlab("Raw expression counts") +
  ylab("Number of genes")
```

```{r}
png("./results/count distribution.png", res=300, height=1800, width=1800)
ggplot(benefitialinnoculum.rawCount) +
  geom_histogram(aes(x = Sl1), stat = "bin", bins = 200) +
  xlab("Raw expression counts") +
  ylab("Number of genes")
dev.off()
```
```{r}
write.csv(benefitialinnoculum.rawCount, file="./results/benefitialinnoculum.rawCounts.csv")
```

```{r}
benefitialinnoculum.normCPM <- cpm(calcNormFactors(benefitialinnoculum))
dim(benefitialinnoculum.normCPM)
```

```{r}
head(benefitialinnoculum.normCPM)
```

```{r}
write.csv(benefitialinnoculum.normCPM, file="./results/benefitialinnoculum.normCPM.csv")
```

#Filter counts per million (CPM) out and get smaller size of libraries


```{r}
benefitialinnoculum.filtered <- rowSums(cpm(benefitialinnoculum)>1) >=3
```

```{r}
table(benefitialinnoculum.filtered)
```

```{r}
benefitialinnoculum$samples$lib.size
```

```{r}
benefitialinnoculum <- benefitialinnoculum[benefitialinnoculum.filtered,]
```

```{r}
colSums(benefitialinnoculum$counts)
```

```{r}
dim(benefitialinnoculum)
```
```{r}
benefitialinnoculum$samples$lib.size <- colSums(benefitialinnoculum$counts)
benefitialinnoculum$samples
```

#Count normalization after filter


```{r}
benefitialinnoculum = calcNormFactors(benefitialinnoculum)
```

```{r}
benefitialinnoculum$samples
```

```{r}
benefitialinnoculum.filtered.normCPM <-cpm(calcNormFactors(benefitialinnoculum))
```

```{r}
write.csv(benefitialinnoculum.filtered.normCPM, file="./results/benefitialinnoculum.filtered.normCPM.csv")
```

#Experimental design Treatment factor
```{r}
group<-factor(c('Sl','Sl','Sl',"Ctr","Ctr","Ctr"))
```


```{r}
benefitialinnoculum.design <- model.matrix(~group)   
rownames(benefitialinnoculum.design)<-colnames(benefitialinnoculum$counts)
benefitialinnoculum.design
```

#Sample-level quality control by multidimensional scaling (MDS)

```{r}
plotMDS(benefitialinnoculum, main="MDS plot of RNA-Seq", labels=colnames(benefitialinnoculum$counts))
```


```{r}
png("./results/plotMDS.benefitialinnoculum.png", res=300, height=1800, width=1800)
plotMDS(benefitialinnoculum, main="MDS plot of benefitialinnoculum RNA-Seq", labels=colnames(benefitialinnoculum$counts))
dev.off()
```

#Estimating the common dispersion, trended dispersion, tagwwise dispersion

```{r}
benefitialinnoculum <- estimateGLMCommonDisp(benefitialinnoculum, benefitialinnoculum.design)
```

```{r}
benefitialinnoculum <- estimateGLMTrendedDisp(benefitialinnoculum, benefitialinnoculum.design)
```

```{r}
benefitialinnoculum <- estimateGLMTagwiseDisp(benefitialinnoculum, benefitialinnoculum.design)
```

```{r}
plotMeanVar(benefitialinnoculum, show.tagwise.vars=T,NBline=T)
```

```{r}
plotBCV(benefitialinnoculum)
```

#Fit DGEList and design matrix to genewise GLM


```{r}
benefitialinnoculum.fit <- glmFit(benefitialinnoculum, benefitialinnoculum.design)
colnames(benefitialinnoculum.fit)
```

```{r}
lrt.Sl_vs_Ctr <- glmLRT(benefitialinnoculum.fit, coef=2) 
```

```{r}
t1<-topTags(lrt.Sl_vs_Ctr, n=nrow(benefitialinnoculum))
head(t1$table)
```

#Extract number of differentially expressed (DE) genes

```{r}
summary(decideTests(lrt.Sl_vs_Ctr, adjust.method="BH", p.value=0.05))
```

```{r}
nrow(subset(topTags(lrt.Sl_vs_Ctr, n=586)$table,  logFC > 0))
```

```{r}
lrt.Sl_vs_Ctr_UP <- subset(topTags(lrt.Sl_vs_Ctr, n=586)$table, logFC > 0)
```

```{r}
nrow(subset(topTags(lrt.Sl_vs_Ctr, n=586)$table,  logFC < 0))
```

```{r}
lrt.Sl_vs_Ctr_DW <- subset(topTags(lrt.Sl_vs_Ctr, n=586)$table, logFC < 0)
```



#Differentially expressed transcripts’ tag


```{r}
DEtags.lrt.Sl_vs_Ctr <- rownames(benefitialinnoculum)[as.logical(decideTests(lrt.Sl_vs_Ctr, adjust.method="BH", p.value=0.05))]
```

```{r}
write.csv(lrt.Sl_vs_Ctr_UP, file="./results/lrt.Sl_vs_Ctr_UP.csv")
write.csv(lrt.Sl_vs_Ctr_DW, file="./results/lrt.Sl_vs_Ctr_DW.csv")
```

```{r}
benefitialinnoculum.colSlvsCtr = rep('grey55', nrow(benefitialinnoculum))
```

```{r}
benefitialinnoculum.colSlvsCtr[lrt.Sl_vs_Ctr$table$PValue < 0.05 & lrt.Sl_vs_Ctr$table$logFC >0 ] <- "red"
benefitialinnoculum.colSlvsCtr[lrt.Sl_vs_Ctr$table$PValue < 0.05 & lrt.Sl_vs_Ctr$table$logFC <0 ] <- "blue"

```

```{r}
par(omi=c(0.1,0.1,0.1,0.1), las=1, cex=0.5, mgp=c(3,1,0), cex.main=1.8, cex.lab=1.4, cex.axis=1.4)
plotSmear(lrt.Sl_vs_Ctr, de.tags=DEtags.lrt.Sl_vs_Ctr, xlab="log-counts per million (logCPM)", ylab="log2-fold change (log2FC)", main="Sl benefitialinnoculum compared to Control", pch=19, cex=0.4, smearWidth=0.5, panel.first=grid(), smooth.scatter=FALSE, ylim=c(-7,7), yaxs="i")
```

```{r}
par(omi=c(0.1,0.1,0.1,0.1), las=1, cex=0.5, mgp=c(3,1,0), cex.main=1.8, cex.lab=1.4, cex.axis=1.4)
plotSmear(lrt.Sl_vs_Ctr, xlab="log-counts per million (logCPM)", ylab="log2-fold change (log2FC)", main="a benefitialinnoculum compared to Control", smearWidth=0.5, pch=21, cex=0.4, deCol="red", col=benefitialinnoculum.colSlvsCtr, ylim=c(-7,7), yaxs="i")
abline(h=c(-1,1),col="dodgerblue")
```


```{r}
abline(h=c(-1,1),col="dodgerblue")
```



```{r}
png("./results/plotSmear.benefitialinnoculumRNAseq.png", res=300, height=1800, width=1800)
par(omi=c(0.1,0.1,0.1,0.1), las=1, cex=0.5, mgp=c(3,1,0), cex.main=1.8, cex.lab=1.4, cex.axis=1.4)
plotSmear(lrt.Sl_vs_Ctr, xlab="log-counts per million (logCPM)", ylab="log2-fold change (log2FC)", main="Sl benefitialinnoculum compared to Control", smearWidth=0.5, pch=21, cex=0.4, deCol="red", col=benefitialinnoculum.colSlvsCtr, ylim=c(-7,7), yaxs="i")
abline(h=c(-1,1),col="dodgerblue")
dev.off()
```

